---
title: "Sprague River Basin - WQ Dataset Summary"
author: "Jeffrey D Walker, PhD"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    css: styles.css
    number_sections: yes
    toc: yes
    toc_depth: 4
---

# Overview

This document summarizes the flow and water quality data for Sprague River basin.

```{r libraries, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
library(lubridate)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gridExtra)
theme_set(theme_bw())

opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```

----

# Klamath Tribes - Sprague River Dataset

```{r load kt_sprague}
load('../kt_sprague.Rdata')

q.kt_sprague <- filter(wq.kt_sprague$RAW, VAR=="FLOW")

# add flow and stage columns
wq.kt_sprague <- select(wq.kt_sprague$RAW, SITE, SITE_NAME, DATE, VAR, VALUE) %>%
  spread(VAR, VALUE) %>%
  select(SITE, SITE_NAME, DATE, FLOW, STAGE, TP, PO4, TN, NO23, NH4, TURBIDITY, TSS) %>%
  gather(VAR, VALUE, TP:TSS) %>%
  filter(!is.na(VALUE)) %>%
  arrange(SITE, DATE)
```

## Sample Dates

This figure shows the number of sample dates for each site by year. The dates with and without flow are also indicated.

```{r sample count flow, fig.width=8, fig.height=12}
p.yr <- select(wq.kt_sprague, SITE_NAME, DATE, FLOW) %>%
  unique() %>%
  mutate(HAS_FLOW=ifelse(!is.na(FLOW), 'Flow', 'No Flow')) %>%
  group_by(SITE_NAME, YEAR=year(DATE), HAS_FLOW) %>%
  summarise(N=n()) %>%
  ggplot(aes(YEAR, N, fill=HAS_FLOW)) +
  geom_bar(stat='identity', position='stack') +
  facet_wrap(~SITE_NAME, ncol=1) +
  scale_fill_manual('', values=c('Flow'='steelblue', 'No Flow'='orangered')) +
  labs(x='Year', y='No. Sample Dates') +
  theme(legend.position='top')
p.ts <- select(wq.kt_sprague, SITE_NAME, DATE, FLOW) %>%
  unique() %>%
  mutate(HAS_FLOW=ifelse(!is.na(FLOW), 'Flow', 'No Flow')) %>%
  ggplot(aes(DATE, HAS_FLOW, color=HAS_FLOW)) +
  geom_point() +
  facet_wrap(~SITE_NAME, ncol=1) +
  scale_color_manual('', values=c('Flow'='steelblue', 'No Flow'='orangered')) +
  labs(x='', y='') +
  theme(legend.position='top')
grid.arrange(p.yr, p.ts, nrow=1)
```

This figure also shows the number of samples for each site and variable by month and year. There are some months with no data (Oct 2004, Oct 2010, Oct 2011, June 2013, and Dec 2013). 

```{r cnt var year site tile, fig.width=16, fig.height=10}
group_by(wq.kt_sprague, SITE_NAME, VAR, YEAR=year(DATE), MONTH=month(DATE)) %>%
  summarise(N=n()) %>%
  ggplot(aes(factor(MONTH), SITE_NAME, fill=ordered(N, levels=seq(5,1)))) +
  geom_tile() +
  facet_grid(VAR~YEAR) +
  labs(x="Month", y="Site") +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5)) +
  scale_fill_brewer('No. Samples', palette='Spectral')
```

This figure shows the same data but grouped by site and then variable. This is useful for identifying missing variables.

```{r cnt var year site tile 2, fig.width=16, fig.height=10}
group_by(wq.kt_sprague, SITE_NAME, VAR, YEAR=year(DATE), MONTH=month(DATE)) %>%
  summarise(N=n()) %>%
  ggplot(aes(factor(MONTH), VAR, fill=ordered(N, levels=seq(5,1)))) +
  geom_tile() +
  facet_grid(SITE_NAME~YEAR) +
  labs(x="Month", y="Variable") +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5)) +
  scale_fill_brewer('No. Samples', palette='Spectral')
```

# Klamath Tribes - Tributary Inflow Dataset

The K-T tributary inflow dataset includes one station near the outlet of the Sprague River at Kircher's Bridge.

```{r load kt_trib}
load('../../kt_trib.Rdata')
wq.kt_trib <- filter(wq.kt_trib, 
                     SITE=="WR1000", 
                     VAR %in% c('NH4', 'NO23', 'PO4', 'TN', 'TP')) %>% droplevels
```

```{r plot kt_trib wq ts, fig.width=8, fig.height=8}
p.lin <- ggplot(wq.kt_trib, aes(DATE, VALUE)) +
  geom_point(alpha=0.5) +
  facet_wrap(~VAR, scales='free_y', ncol=1) +
  labs(x="", y="Concentration (ppm)") +
  ggtitle('Linear Scale')
p.log <- ggplot(wq.kt_trib, aes(DATE, VALUE)) +
  geom_point(alpha=0.5) +
  scale_y_log10() +
  ggtitle('Log10 Scale') +
  labs(x="", y="Concentration (ppm)") +
  facet_wrap(~VAR, scales='free_y', ncol=1)
grid.arrange(p.lin, p.log, ncol=2)
```

## Compare Kircher's to Power

The Kircher station is located about 7 miles downstream of the Power station. The Chiloquin dam was between the two stations until summer 2008 when it was removed.

```{r wq outlet}
wq.outlet <- rbind(mutate(wq.kt_trib, SITE_NAME="Sprague_Kirchers") %>% 
                     select(SITE, SITE_NAME, DATE, VAR, VALUE),
                   filter(wq.kt_sprague, SITE_NAME=="Power") %>% 
                     select(SITE, SITE_NAME, DATE, VAR, VALUE)) %>%
  droplevels %>%
  mutate(SITE=factor(as.character(SITE)),
         SITE_NAME=factor(as.character(SITE_NAME)))
```

This figure shows the time series for each water quality variable on linear and log scales.

```{r plot outlet wq ts, fig.width=8, fig.height=10}
p.lin <- wq.outlet %>%
  ggplot(aes(DATE, VALUE, color=SITE_NAME)) +
  geom_point(size=1) +
  facet_wrap(~VAR, ncol=1, scales='free_y') +
  scale_color_manual('Site', values=c('Sprague_Kirchers'='steelblue', 'Power'='orangered')) +
  theme(legend.position='top') +
  ggtitle('Linear Scale') +
  labs(x="", y="Concentration (ppm)")
p.log <- wq.outlet %>%
  ggplot(aes(DATE, VALUE, color=SITE_NAME)) +
  geom_point(size=1) +
  scale_y_log10() +
  facet_wrap(~VAR, ncol=1, scales='free_y') +
  scale_color_manual('Site', values=c('Sprague_Kirchers'='steelblue', 'Power'='orangered')) +
  theme(legend.position='top') +
  ggtitle('Log Scale') +
  labs(x="", y="Concentration (ppm)")
grid.arrange(p.lin, p.log, ncol=2)
```

This figure shows the same data as boxplots grouped by year, variable, and site on a log scale.

```{r plot outlet wq box, fig.width=7, fig.height=8}
wq.outlet %>%
  mutate(YEAR=year(DATE)) %>%
  filter(YEAR %in% seq(2001, 2013), VAR %in% c("TP", "PO4", "TN", "NO23", "NH4")) %>%
  ggplot(aes(factor(YEAR), VALUE, fill=SITE_NAME)) +
  geom_boxplot() +
  scale_y_log10() +
  facet_wrap(~VAR, ncol=1, scales='free_y') +
  scale_fill_manual('Site', values=c('Sprague_Kirchers'='steelblue', 'Power'='orangered')) +
  labs(x="Year", y="Concentration (ppm)") +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
        legend.position='top')
```
