---
title: "Sprague River Basin - WQ Sampling Frequency"
author: "Jeffrey D Walker, PhD"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    css: styles.css
---

This document summarizes the frequency of sampling for Klamath Tribes Sprague River dataset.

```{r load, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
library(lubridate)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gridExtra)
theme_set(theme_bw())

opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```


```{r load kt_sprague}
load('../kt_sprague.Rdata')

q.kt_sprague <- filter(wq.kt_sprague$RAW, VAR=="FLOW")

# add flow and stage columns
wq.kt_sprague <- select(wq.kt_sprague$RAW, -UNITS, -DATETIME) %>%
  spread(VAR, VALUE) %>%
  select(SITE, SITE_NAME, DATE, FLOW, STAGE, TP, PO4, TN, NO23, NH4, TURBIDITY, TSS) %>%
  gather(VAR, VALUE, TP:TSS) %>%
  filter(!is.na(VALUE)) %>%
  arrange(SITE_NAME, DATE)
```

The sampling program targets biweekly sampling intervals. This figure shows a histogram of the number of days between samples for each site. The peaks occur around 14 days (indicated by the red line), but there are also longer periods due to skipped sample dates, and shorter periods.

```{r date diff hist, fig.width=8, fig.height=4}
select(wq.kt_sprague, SITE_NAME, DATE) %>%
  unique() %>%
  group_by(SITE_NAME) %>%
  mutate(DATE_LAG=lag(DATE),
         DATE_DIFF=difftime(DATE,DATE_LAG,units='day') %>% as.numeric) %>%
  arrange(DATE_DIFF) %>%
  filter(DATE_DIFF < 1000) %>%
  ggplot(aes(DATE_DIFF)) +
  geom_histogram(binwidth=1) +
  geom_vline(xint=14, color='red', linetype=2) +
  facet_wrap(~SITE_NAME, ncol=4) +
  labs(x="Days between Samples", y="Number of Sample Dates")
```

Here is the same figure excluding dates with more than 30 days from the previous date (smaller x-axis scale).

```{r date diff hist zoom, fig.width=8, fig.height=4}
select(wq.kt_sprague, SITE_NAME, DATE) %>%
  unique() %>%
  group_by(SITE_NAME) %>%
  mutate(DATE_LAG=lag(DATE),
         DATE_DIFF=difftime(DATE,DATE_LAG,units='day') %>% as.numeric) %>%
  arrange(DATE_DIFF) %>%
  filter(DATE_DIFF < 30) %>%
  ggplot(aes(DATE_DIFF)) +
  geom_histogram(binwidth=1) +
  geom_vline(xint=14, color='red', linetype=2) +
  facet_wrap(~SITE_NAME, ncol=4) +
  labs(x="Days between Samples", y="Number of Sample Dates")
```

This figure shows the days since the previous sample as a time series. The deviations from 14 days occurred more or less randomly over the period of record.

```{r date diff ts, fig.width=8, fig.height=4}
filter(wq.kt_sprague, VAR=="TP") %>%
  select(SITE_NAME, DATE) %>%
  unique() %>%
  group_by(SITE_NAME) %>%
  mutate(DATE_LAG=lag(DATE),
         DATE_DIFF=difftime(DATE,DATE_LAG,units='day') %>% as.numeric) %>%
  arrange(DATE_DIFF) %>%
  filter(DATE_DIFF<1000) %>%
  ggplot(aes(DATE_LAG, DATE_DIFF)) +
  geom_point(size=1) +
  geom_hline(yint=14, color='red', linetype=2) +
  facet_wrap(~SITE_NAME, ncol=4) +
  labs(x='', y='Days since last sample') +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
```

This figure shows the absolute relative difference in successive TP measurements vs. days between the samples. One would expect smaller differences corresponding to fewer days between samples, but this figure shows some pairs of samples separated by only one or a couple of days that are largely different. The absolute relative difference is absolute value of the difference between a pair of samples divided by the mean of the two samples.

```{r tp abs rel diff, fig.width=8, fig.height=4}
filter(wq.kt_sprague, VAR=="TP") %>%
  group_by(SITE_NAME) %>%
  mutate(DATE_LAG=lag(DATE),
         VALUE_LAG=lag(VALUE),
         VALUE_DIFF=VALUE_LAG-VALUE,
         VALUE_DIFF_REL=VALUE_DIFF/((VALUE+VALUE_LAG)/2),
         VALUE_DIFF_REL_ABS=abs(VALUE_DIFF_REL),
         FLOW_LAG=lag(FLOW),
         FLOW_DIFF=FLOW_LAG-FLOW,
         FLOW_DIFF_REL=FLOW_DIFF/((FLOW+FLOW_LAG)/2),
         FLOW_DIFF_REL_ABS=abs(FLOW_DIFF_REL),
         DATE_DIFF=difftime(DATE,DATE_LAG,units='day') %>% as.numeric) %>%
  filter(DATE_DIFF<30) %>%
  ggplot(aes(DATE_DIFF, VALUE_DIFF_REL_ABS)) +
  geom_point(alpha=0.5) +
  facet_wrap(~SITE_NAME, ncol=4) +
  labs(x='Days between Samples', y='Absolute Relative Difference in TP(t), TP(t+1)')
```

