---
title: "Trend Periods"
author: "Jeffrey D Walker, PhD"
date: "January 11, 2016"
output: html_document
---
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(knitr)
library(ggmap)
library(gridExtra)
library(scales)
library(fluxr)
library(wq)
library(zoo)
theme_set(theme_bw())
```


```{r, cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
term_labs <- c('Q'='Flow', 'L'='Load', 'C'='FWM Conc')
term_units <- c('C'='ppb', 'L'='kg/d', 'Q'='hm3/d')
dataset <- "POR"
load('../loads.Rdata')
load('../kt_sprague.Rdata')
load('../prism.Rdata')

source('../functions.R')

# load data ----
year_range <- 2002:2014

df_wyr <- loads_df[['wyr']] %>%
  filter(DATASET=="POR",
         TERM %in% c('Q', 'L', 'C'),
         VAR != "TSS",
         SEASON=="Annual",
         SITE_NAME %in% as.character(stn.kt_sprague$SITE_NAME),
         !(SITE_NAME %in% c('SF_Ivory', 'NF_Ivory'))) %>%
  select(-DATASET) %>%
  spread(TERM, VALUE) %>%
  droplevels

df_wyr.flow <- filter(df_wyr, VAR=='FLOW') %>%
  select(SITE_NAME, WYEAR, Q)
df_wyr <- filter(df_wyr, VAR!='FLOW') %>%
  select(-Q) %>%
  left_join(df_wyr.flow, by=c('SITE_NAME', 'WYEAR'))

df_mon <- loads_df[['mon']] %>%
  filter(DATASET=="POR",
         TERM %in% c('Q', 'L', 'C'),
         VAR != "TSS",
         SITE_NAME %in% as.character(stn.kt_sprague$SITE_NAME),
         !(SITE_NAME %in% c('SF_Ivory', 'NF_Ivory'))) %>%
  select(-DATASET) %>%
  mutate(N_DAY=days_in_month(MONTHYEAR)) %>%
  spread(TERM, VALUE) %>%
  rename(DATE=MONTHYEAR) %>%
  droplevels

df_mon.flow <- filter(df_mon, VAR=='FLOW') %>%
  select(SITE_NAME, DATE, Q)
df_mon <- filter(df_mon, VAR!='FLOW') %>%
  select(-Q) %>%
  left_join(df_mon.flow, by=c('SITE_NAME', 'DATE'))

# trend functions ----
trend.sk <- function(x, value_var, months, month_label, years, log_trans=TRUE, water_year=TRUE) {
  # months <- 1:4
  # month_label <- 'Annual'
  # years <- 1992:2010
  # water_year <- TRUE

  if (water_year==TRUE) {
    x <- subset(x, MONTH %in% months & WYEAR %in% years)
  } else {
    x <- subset(x, MONTH %in% months & YEAR %in% years)
  }
  if (log_trans) {
    z <- zoo(log10(x[[value_var]]), x[['DATE']])
  } else {
    z <- zoo(x[[value_var]], x[['DATE']])
  }

  z <- aggregate(z, as.yearmon, mean, na.rm=TRUE)
  z <- as.ts(z)
  z <- window(z, start=c(start(z)[1],1), end=c(end(z)[1], frequency(z)), extend=TRUE)

  if (length(months)>1) {
    t <- z[cycle(z) %in% months]
    t <- ts(t, start=c(start(z)[1],1), end=c(end(z)[1],length(months)), frequency=length(months))
  } else {
    t <- z
  }

  sk <- seaKen(t)
#   slope <- ifelse(abs(sk$sen.slope) < 1E-5, 0, sk$sen.slope)
  slope <- sk$sen.slope
  if (log_trans) {
    slope.pct <- slope.pct <- 10^slope-1
  } else {
    slope.pct <- sk$sen.slope.pct/100
  }

  intercept <- median(t - slope*time(t), na.rm=T)
  pval <- sk$p.value
  sig <- cut(pval, breaks=c(0,0.05,0.1,1), labels=c("p<0.05","0.05<p<0.10","p>0.10"))
  if (is.na(sig)) sig <- "p>0.10"

  return(data.frame(TERM=value_var,
                    LOG=log_trans,
                    YEAR_SPAN=paste(min(years),max(years),sep='-'),
                    MONTH_LABEL=month_label,
                    METHOD='SeasonalKendall',
                    MEAN.VAL=mean(t, na.rm=T),
                    MEAN.TIME=decimal_date(mean(x$DATE)),
                    INTERCEPT=intercept,
                    SLOPE=slope,
                    SLOPE.PCT=slope.pct,
                    PVAL=pval,
                    SIGNIF=sig,
                    DIRECTION=ordered(ifelse(slope>0, 'Increasing', 'Decreasing'), levels=c('Increasing', 'Decreasing')))		)
}
# trend.sk(x=filter(df_mon, SITE_NAME=="Power", VAR=="TP"),
#          value_var='C',
#          months=1:12,
#          month_label='Seasonal',
#          years=2002:2014,
#          log_trans=TRUE,
#          water_year=TRUE)
# trend.sk(x=filter(df_mon, SITE_NAME=="Power", VAR=="TP"),
#          value_var='C',
#          months=10,
#          month_label='10',
#          years=2002:2014,
#          log_trans=TRUE,
#          water_year=TRUE)

trend.mk <- function(x, value_var, years, month_label, log_trans=FALSE, water_year=TRUE) {
  if (water_year==TRUE) {
    x <- subset(x, WYEAR %in% years)
    start_yr <- min(x[['WYEAR']])
  } else {
    x <- subset(x, YEAR %in% years)
    start_yr <- min(x[['YEAR']])
  }

  if (log_trans) {
    z <- ts(log10(x[[value_var]]), start=start_yr, freq=1)
  } else {
    z <- ts(x[[value_var]], start=start_yr, freq=1)
  }

  mk <- mannKen(z)

#   slope <- ifelse(abs(mk$sen.slope) < 1E-5, 0, mk$sen.slope)
#   slope.pct <- mk$sen.slope.pct/100
#
  slope <- mk$sen.slope
  if (log_trans) {
    slope.pct <- 10^slope-1
  } else {
    slope.pct <- mk$sen.slope.pct/100
  }

  intercept <- median(z - slope*time(z), na.rm=T)
  pval <- mk$p.value
  sig <- cut(pval, breaks=c(0,0.05,0.1,1), labels=c("p<0.05","0.05<p<0.10","p>0.10"))
  if (is.na(sig)) sig <- "p>0.10"

  data.frame(TERM=value_var,
             LOG=log_trans,
             YEAR_SPAN=paste(min(years),max(years),sep='-'),
             MONTH_LABEL=month_label,
             METHOD='MannKendall',
             MEAN.VAL=mean(z, na.rm=T),
             MEAN.TIME=mean(time(z)),
             INTERCEPT=intercept,
             SLOPE=slope,
             SLOPE.PCT=slope.pct,
             PVAL=pval,
             SIGNIF=sig,
             DIRECTION=ordered(ifelse(slope>0, 'Increasing', 'Decreasing'), levels=c('Increasing', 'Decreasing')))
}
# trend.mk(x=filter(df_wyr, SITE_NAME=="Power", VAR=="TP"),
#          value_var='C',
#          month_label='Annual',
#          years=2002:2014,
#          log_trans=TRUE,
#          water_year=TRUE)

trend.lm <- function(x, value_var, years, month_label, log_trans=FALSE, water_year=TRUE) {
  if (water_year==TRUE) {
    x <- subset(x, WYEAR %in% years)
    mean_time <- mean(x[['WYEAR']])
    form <- as.formula(paste(value_var, 'WYEAR', sep=' ~ '))
  } else {
    x <- subset(x, YEAR %in% years)
    mean_time <- mean(x[['YEAR']])
    form <- as.formula(paste(value_var, 'YEAR', sep=' ~ '))
  }

  if (log_trans) {
    x[[value_var]] <- log10(x[[value_var]])
  }

  fit <- lm(form, data=x)

  intercept <- coef(fit)[[1]]
  slope <- coef(fit)[[2]]
  if (log_trans) {
    slope.pct <- 10^slope-1
  } else {
    slope.pct <- slope/mean(x[[value_var]])
  }

  pval <- summary(fit)$coeff['WYEAR', 'Pr(>|t|)']
  sig <- cut(pval, breaks=c(0,0.05,0.1,1), labels=c("p<0.05","0.05<p<0.10","p>0.10"))
  if (is.na(sig)) sig <- "p>0.10"

  data.frame(TERM=value_var,
             LOG=log_trans,
             YEAR_SPAN=paste(min(years),max(years),sep='-'),
             MONTH_LABEL=month_label,
             METHOD='LinearRegression',
             MEAN.VAL=mean(x[[value_var]]),
             MEAN.TIME=mean_time,
             INTERCEPT=intercept,
             SLOPE=slope,
             SLOPE.PCT=slope.pct,
             PVAL=pval,
             SIGNIF=sig,
             DIRECTION=ordered(ifelse(slope>0, 'Increasing', 'Decreasing'), levels=c('Increasing', 'Decreasing')))
}
# trend.lm(x=filter(df_wyr, SITE_NAME=="Power", VAR=="TP"),
#          value_var='C',
#          month_label='Annual',
#          years=2002:2014,
#          log_trans=TRUE,
#          water_year=TRUE)

trend.batch <- function(x_mon, x_wyr, years, log_trans=FALSE, water_year=TRUE) {
  x.site_name <- unique(x_mon$SITE_NAME)
  x.var <- unique(x_mon$VAR)
  stopifnot(length(x.site_name)==1)
  stopifnot(length(x.var)==1)

  batch.mon <- function(x, value_var) {
    df.seasonal <- trend.sk(x, value_var=value_var,
                            months=1:12, month_label='All Months',
                            years=years, log_trans=log_trans, water_year=water_year)
    df.4_9 <- trend.sk(x, value_var=value_var,
                          months=4:9, month_label='Apr-Sep',
                          years=years, log_trans=log_trans, water_year=water_year)
    df.10_3 <- trend.sk(x, value_var=value_var,
                          months=c(1:3, 10:12), month_label='Oct-Mar',
                          years=years, log_trans=log_trans, water_year=water_year)
    df.fall <- trend.sk(x, value_var=value_var,
                          months=c(10:12), month_label='Oct-Dec',
                          years=years, log_trans=log_trans, water_year=water_year)
    df.winter <- trend.sk(x, value_var=value_var,
                        months=1:3, month_label='Jan-Mar',
                        years=years, log_trans=log_trans, water_year=water_year)
    df.spring <- trend.sk(x, value_var=value_var,
                          months=4:6, month_label='Apr-Jun',
                          years=years, log_trans=log_trans, water_year=water_year)
    df.summer <- trend.sk(x, value_var=value_var,
                          months=7:9, month_label='Jul-Sep',
                          years=years, log_trans=log_trans, water_year=water_year)
    df <- rbind(df.seasonal, df.fall, df.winter, df.spring, df.summer, df.4_9, df.10_3)

    for (m in 1:12) {
      df.m <- trend.sk(x=x, value_var=value_var,
                       months=m, month_label=as.character(m),
                       years=years, log_trans=log_trans, water_year=water_year)
      df <- rbind(df, df.m)
    }

    df
  }

  t_mon_Q <- batch.mon(x_mon, value_var='Q')
  t_mon_L <- batch.mon(x_mon, value_var='L')
  t_mon_C <- batch.mon(x_mon, value_var='C')
  t_mon <- rbind(t_mon_Q, t_mon_L, t_mon_C)

  t_wyr_mk_Q <- trend.mk(x_wyr, value_var='Q', years=years, log_trans=log_trans, month_label='Annual-MK')
  t_wyr_mk_L <- trend.mk(x_wyr, value_var='L', years=years, log_trans=log_trans, month_label='Annual-MK')
  t_wyr_mk_C <- trend.mk(x_wyr, value_var='C', years=years, log_trans=log_trans, month_label='Annual-MK')
  t_wyr_mk <- rbind(t_wyr_mk_Q, t_wyr_mk_L, t_wyr_mk_C)

  t_wyr_lm_Q <- trend.lm(x_wyr, value_var='Q', years=years, log_trans=log_trans, month_label='Annual-Reg')
  t_wyr_lm_L <- trend.lm(x_wyr, value_var='L', years=years, log_trans=log_trans, month_label='Annual-Reg')
  t_wyr_lm_C <- trend.lm(x_wyr, value_var='C', years=years, log_trans=log_trans, month_label='Annual-Reg')
  t_wyr_lm <- rbind(t_wyr_lm_Q, t_wyr_lm_L, t_wyr_lm_C)

  t_wyr <- rbind(t_wyr_mk, t_wyr_lm)

  t_all <- rbind(t_mon, t_wyr)

  t_all <- mutate(t_all,
                  MONTH_LABEL=ordered(MONTH_LABEL,
                                      levels=c(as.character(10:12), as.character(1:9),
                                               'Oct-Dec','Jan-Mar','Apr-Jun','Jul-Sep',
                                               'Oct-Mar','Apr-Sep','All Months',
                                               'Annual-MK','Annual-Reg')),
                  TERM=ordered(TERM, levels=names(term_labs)))
  t_all
}
# trend.batch(x_mon=filter(df_mon, SITE_NAME=="Power", VAR=="TP"),
#             x_wyr=filter(df_wyr, SITE_NAME=="Power", VAR=="TP"),
#             years=2002:2014,
#             log_trans=TRUE,
#             water_year=TRUE)

```

```{r, cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
# trend.batch(x_mon=filter(df_mon, SITE_NAME=="Power", VAR=="TP"),
#             x_wyr=filter(df_wyr, SITE_NAME=="Power", VAR=="TP"),
#             years=2002:2014,
#             log_trans=TRUE,
#             water_year=TRUE)
# 
# for (start_year in seq(2002, 2013)) {
#   for (end_year in seq(start_year+1, 2014)) {
#     cat(start_year, end_year, '\n')
#   }
# }

# grid.arrange()

trends <- lapply(seq(2002, 2013), function (start_year) {
  # cat('start_year ', start_year, '\n')
  lapply(seq(start_year+1, 2014), function (end_year) {
    # cat('end_year ', end_year, '\n')
    # lapply(as.character(unique(df_mon$VAR)), function(variable) {
    lapply("TP", function(variable) {
      # cat('..', variable, '\n')
      lapply(as.character(levels(df_mon$SITE_NAME)), function(site) {
      # lapply(c("Power"), function(site) {
        # cat('....', site, '\n')
        year_range <- start_year:end_year
        x <- trend.batch(x_mon=filter(df_mon, SITE_NAME==site, VAR==variable),
                         x_wyr=filter(df_wyr, SITE_NAME==site, VAR==variable),
                         years=year_range,
                         log_trans=TRUE,
                         water_year=TRUE)
        x$SITE_NAME <- site
        x$VAR <- variable
        x$START_YEAR <- start_year
        x$END_YEAR <- end_year
        x
      }) %>%
        rbind_all
    }) %>%
      rbind_all
  }) %>%
    rbind_all
}) %>%
  rbind_all

trends2 <- trends %>%
  mutate(SITE_NAME=ordered(SITE_NAME, levels=levels(stn.kt_sprague$SITE_NAME)),
         SIGNIF=ordered(as.character(SIGNIF), levels=c("p<0.05","0.05<p<0.10","p>0.10"))) %>%
  droplevels %>%
  mutate(LABEL=ifelse(SIGNIF=="p>0.10", "Not Significant",
                      ifelse(SIGNIF=="p<0.05",
                             ifelse(DIRECTION=="Increasing",
                                    "Increasing (p<0.05)",
                                    "Decreasing (p<0.05)"),
                             ifelse(DIRECTION=="Increasing",
                                    "Increasing (0.05<p<0.1)",
                                    "Decreasing (0.05<p<0.1)"))),
         LABEL=ordered(LABEL, levels=c("Increasing (p<0.05)",
                                       "Increasing (0.05<p<0.1)",
                                       "Not Significant",
                                       "Decreasing (0.05<p<0.1)",
                                       "Decreasing (p<0.05)")))
```

This figure shows the trend test results for the FWM TP concentration over different annual periods.

The seasonal kendall test was run over all months for each pair of start and end years. The square in the lower-left of each panel shows the result for 2002-2003. The lower-right square is for 2002-2014. The variation and clustering of similar results indicates the importance of the trend analysis period, and also shows the difference between sites. For example, SF and NF show significant increasing trends when the start year ranges from 2002 to ~2009 and the end year ranges from 2011-2014. It also shows that the significant trends from 2002-2014 for Power and Lone_Pine are not significant with one less year of data (2002-2013).

```{r, echo=FALSE, fig.width=10, fig.height=6}
filter(trends2,
       TERM=="C",
       MONTH_LABEL=="All Months") %>%
  ggplot(aes(END_YEAR, START_YEAR, fill=LABEL)) +
  geom_tile(color='black') +
  scale_x_continuous(breaks=pretty_breaks()) +
  scale_y_continuous(breaks=pretty_breaks()) +
  scale_fill_manual('',
                    values=c("Increasing (p<0.05)"='orangered',
                             "Increasing (0.05<p<0.1)"="#FFA895",
                             "Not Significant"='white',
                             "Decreasing (0.05<p<0.1)"="#AAC4DB",
                             "Decreasing (p<0.05)"="steelblue"),
                    drop=FALSE) +
  labs(x="End WYear", y="Start WYear") +
  ggtitle("Trend Results for Varying Periods (Start Year -- End Year)\nVariable: TP, Term: C") +
  facet_wrap(~SITE_NAME, scales="free", nrow=2) +
  theme(panel.grid=element_blank(),
        aspect.ratio=1,
        axis.text.x=element_text(angle=90, hjust=0, vjust=0.5))
```
